---
title: "Cetaceans"
author: "Julianna Renzi"
date: "2/17/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(raster)
library(sf)
library(fasterize) # use sf geometries and rasterize them efficiently
library(rgdal)
library(rnaturalearth) # get coastline of California
library(rnaturalearthdata)
library(rgeos)
library(ggsn)
```

## Read in the data

Read in raster data on California cetaceans, taken from AquaMaps (Kaschner, K., Rius-Barile, J., Kesner-Reyes, K., Garilao, C., Kullander, S., Rees, T., & Froese, R. (2016). AquaMaps: Predicted range maps for aquatic species. www.aquamaps.org). The data cover 35 cetacean species and the probability of seeing each species at a given location given environmental factors such as depth, temperature, salinity, and proximity to shore.

```{r}
# get the names of rasters for each species
cet_tifs <- list.files(path = here("data", "ca_cetaceans"), pattern = ".tif", full.names = TRUE)

# bring rasters in as a raster stack
cet_stack <- raster::stack(cet_tifs)

# can view the object to make sure it looks correct:
cet_stack # looks good

# plot one to get an idea of what the data look like
plot(cet_stack$Balaenoptera_acutorostrata, col = hcl.colors(n = 100))
plot(cet_stack$Globicephala_macrorhynchus, col = hcl.colors(n = 100))
plot(cet_stack$Kogia_sima, col = hcl.colors(n = 100))
```

## Turn the data into presence/absence using raster algebra

For the purposes of this exercise, we'll define a species as likely to be "present" when the probability of observing it is greater than 0.75. 

```{r}
convert_pa <- function(x) {
  ifelse(x > 0.75, 1, 0)
}

cet_presence <- calc(cet_stack, function(x){convert_pa(x)})

# plot to check and make sure it worked
plot(cet_presence$Balaenoptera_acutorostrata, col = hcl.colors(n = 2))
plot(cet_presence$Globicephala_macrorhynchus, col = hcl.colors(n = 2))
plot(cet_presence$Kogia_sima, col = hcl.colors(n = 2))
# all look reasonable given the raw maps above
```
## Calculate and plot potential species richness

```{r}
cet_sum <- calc(cet_presence, sum) # get the sum of 1's (presence as we defined it) for the number of potential species in a grid cell

# can see what the summary looks like
plot(cet_sum, col = hcl.colors(n = 20))

# get it into data.frame format for ggplot
cet_df <- raster::rasterToPoints(cet_sum) %>% # this is WGS84
  as.data.frame()

# get coastline data for reference
coasts <- ne_coastline(scale = "medium", returnclass = "sf")

# can visualize 
plot(coasts) # this is WGS84

# plot with ggplot()
ggplot(data = coasts) +
  geom_sf() +
  coord_sf(xlim = c(-125, -113), ylim = c(28, 38), expand = FALSE) +
  annotate(geom = "text", x = -121, y = 30, label = "Pacific Ocean", 
     fontface = "italic", color = "grey22", size = 3) +
  annotate(geom = "text", x = -118, y = 36, label = "California", 
           color = "grey22", size = 3) +
  geom_raster(data = cet_df, aes(x = x, y = y, fill = layer), alpha = 0.75) +
  scale_fill_gradient(low = "lightblue", high = "darkred",
                           breaks=seq(from = 1, to = 25, by = 2)) +
  theme_bw() +
  theme(axis.title.x = element_blank(), 
     axis.title.y = element_blank(), 
     panel.border = element_rect(fill = NA)) +
  labs(fill = "Species richness",
       caption = expression(atop("Map of predicted cetacean species richness in the California Bight where a prediction", "probability of 0.75 was considered a presence and below that was an absence")))
```

